Efficient evaluation of threshold queries of derived fields in a numerical simulation database

### 2014-10-07 paper.tex
Rearranged some figures. Cleaned up references.
Some minor fixes.

\begin{abstract}
%Data-intensive architectures and compute clusters have been deployed to store, manage and provide public access to such datasets. 
%The systems built on top of these architectures, sometimes called virtual laboratories, usually do not support the analysis functionality that scientists are 
%interested in out of the box. Moreover, d
In this paper we present a method for the efficient evaluation of threshold queries of derived fields for large numerical 
simulation datasets stored in a cluster of relational databases. 
Such simulations are commonplace in the study of many different physical phenomena. 
The datasets produced by these simulations are in the TB and even PB ranges. 
Data-intensive computations that examine entire time-steps of the simulation data are impractical to perform
locally and can take days or months to iterate over the entire dataset. 
The integrated method for the evaluation of threshold queries that we have developed achieves scalability through data-parallel execution. 
We extend the database system with the introduction of an application-aware cache for query results, building on the concept of semantic caching.
The cache has little overhead and improves query performance by over an order of magnitude for queries that hit the cache.
\end{abstract}

@@

@@

%Why hasn't it been solved before? (Or, what's wrong with previous proposed solutions? How does mine differ?)
Data-intensive architectures and compute clusters built from commodity hardware rely on parallel I/O to multiple disks and high network bandwidth
to achieve high throughput. Such systems have only recently been deployed for the storage of large numerical simulation 
datasets. The virtual laboratories built on these systems make use of relational database system technology to store and manage these large array datasets.
Relational database systems however often do not support
all of the functionality that scientists are interested in out of the box. It is either up to the user to develop more sophisticated analysis routines locally or such
capabilities have to be built-in through user-defined functions or stored procedures. 
Database systems support rollup queries, including top-$k$ queries, but in most cases these
queries apply only simple linear score functions on the attribute values of individual records \cite{Ilyas}.
Additionally, many top-$k$ query evaluation techniques rely on the score functions being monotone in order to perform early pruning.
This is an assumption that we cannot make for the functions used to compute all the different possible derived fields of interest in fluid mechanics.
Even approaches that aim to work with general score functions \cite{Deshpande, Xin} assume that the function operates on the attributes of a single record.
In contrast our approach computes the value of a derived field at each location in the array by examining the data at all neighboring locations, which
are within the kernel of the computation.

@@

The JHTDB provide a variety of data-intensive analysis routines that are executed on the database nodes. These include interpolation, 
differentiation, particle tracking and spatial filtering. These tasks are often data-intensive and in order to leverage the capabilities of the cluster we
have developed data-driven batch processing techniques for their evaluation \cite{KanovSC11, KanovSC12}. What differentiates threshold and
top-$k$ queries from these tasks is the fact that they usually operate on small subsets of the space or a collection of individual target locations, whereas
threshold and top-$k$ queries usually have to examine the entire data volume of a time-step. Furthermore, the data product of threshold and top-$k$ 
queries is much smaller in relation to the amount of data that need to be examined. This fact combined with the fact that subsequent queries can reuse 
previously computed results makes the query results suitable to caching.

@@

{\bf Derived fields computation:} The databases store only the raw field data from the simulation (e.g. velocity, pressure, etc.). However, the threshold
queries of most interest to science users produce all locations where the values of a derived field are above a given threshold. Thus, the
derived field in question has to be computed from the raw data first. For most derived fields of interest, this computation has local support. It has an 
associated localized kernel of computation around each grid node. Therefore, the value of the derived field at each grid node depends on the value of the
stored field at all of the grid locations, which are part of the kernel of computation.

@@

Caching the query results
preserves the computational effort in addition to substantially reducing I/O. The cached data are for the particular derived field that was queried
and not the raw data fields.
Thus, we do not have to derive the requested field from the raw data for 
queries that hit the cache. This results in a substantial improvement in query performance as we only have to scan
a small set of data and do not need to perform any additional computation.

@@

So far we have presented the effectiveness of evaluating threshold queries of derived fields on the database cluster storing the raw simulation data. The 
data-parallel computation of the derived fields allows us to evaluate a threshold query over
an entire $1024^3$ time-step part of a 20TB dataset in less than two minutes. The introduction of an application-aware cache for the query results of these
queries reduces this time to several seconds when there is a cache hit. In contrast, an evaluation of this functionality performed locally by the user takes
over 20 hours. To perform the evaluation locally the user requests the derived field of interest from the database by submitting multiple queries over
subregions of a time-step. This is necessary as requesting a derived field over an entire time-step will overload the network. Derived fields may have even
more components than the scalar or vector field stored in the database. For example, the velocity gradient (needed for the computation of the vorticity) has 9
components compared with the 3 components of the velocity. Given a single-precision floating-point representation, this makes the velocity gradient of an
entire time-step at least 36GB in size. A Web-service request will be much larger due to the overhead of wrapping the data in an xml format. After the
field of interest is obtained locally the user has to threshold it to get the final result, which is reasonably fast, but discards most of the data that have been
requested to yield a small in size result.

\subsection{Execution Time Breakdown}
\kk{OR}
\subsection{Evaluation of Additional Fields}
The data-parallel evaluation of threshold queries shows stable execution time for different derived fields in addition to the different threshold levels and
time-steps queried. The execution times depend on the complexity of the computation needed to evaluate the particular derived field requested. Figures 
\ref{fig:vorticity_breakdown}, \ref{fig:q_breakdown}, \ref{fig:magnetic_breakdown} show a breakdown of the execution time of threshold queries of different
derived fields, which are evaluated from the raw data and a cold cache using 4 processes per node on a 4 node cluster. Almost the entire time is spent
performing the I/O and computation associated with the derived field requested.

The vorticity field and the Q-criterion have similar I/O requirements as they
have the same kernels of computation and both can be derived from the velocity gradient. The vorticity has 3 components and its computation only examines
6 of the 9 components of the velocity gradient, which are also examined in pairs (see Eq. \ref{eq:curl}). On the other hand, even though the Q-criterion is a
scalar value, it is computed through a non-linear combination of all 9 of the components of the velocity gradient. This means that the velocity gradient has
to be computed at each grid location before the Q-criterion is evaluated, which is reflected in the increased computation time that we observer for the
Q-criterion. The magnetic field is one of the raw fields of the magnetohydrodynamics dataset that are stored in the database. Therefore, there is no
additional computation needed to derive it from the data, every data point has to be simply compared with the threshold level specified. This is why the 
computation time is much smaller compared to the queries for the vorticity and the Q-criterion. The I/O time for the magnetic field is also smaller. This is
because its kernel of computation is a single point and therefore there are no additional data along the boundary that have to be requested from adjacent 
nodes. In that case all of the data needed are available locally.

In all of these cases the time take to interrogate the cache is negligible. The mediator time to distribute the queries and assemble the results as well as the
time to transfer them to the user are also substantially smaller than the I/O and computation times. As expected they increase with the number of points in
the result set.

@@

\section{Related Work}
Only select few database systems offer support for arrays as first-class citizens. Even fewer provide the fault-tolerance, scalability and availability guarantees 
necessary for a system managing multi-terabyte datasets in a production setting. This is part of the reason why we have chosen to represent the array simulation
data in the JHTDB as a collection of binary large objects in a standard relational DBMS and perform the array manipulation tasks necessary at the application level.
The systems that provide
support for arrays and aim to handle large array data efficiently are RasDaMan \cite{Rasdaman}, SciDB \cite{Scidb} and MonetDB/SciQL \cite{Sciql}. RasDaMan
partitions raster objects into tiles, which are stored in a traditional relational database system. This approach is similar to how the numerical simulation data is 
handled in the JHTDB. RasDaMan provides RasQL \cite{Rasql}, which is a SQL-92 based query language for the manipulation of raster images. SciDB is 
an array database system build from the ground up. Array attributes are partitioned vertically and each attribute array is decomposed into overlapping chunks.
SciDB provides a declarative Array Query Language (AQL) and an Array Functional Language (AFL). Users can create arrays with named dimensions with
AQL and make use of the functional operators defined in AFL, such
as SLICE, SUBSAMPLE, SJOIN, FILTER and APPLY. SciQL's focus is on language design and the integration with SQL:2003 syntax and semantics. It is implemented
within the MonetDB framework \cite{MonetDB}.

@@

The Web-services approach to archived numerical simulation datasets provides public access to high quality simulation data to anyone with an internet
connection. The Web-services methods can be called from any modern programming language and we provide C, Fortran and Matlab client libraries for the
JHTDB. The evaluation of each query submitted through a Web-service call is carried out on the nodes of the database cluster by mens of a stored procedure
or a user-defined function that has been implemented and deployed to handle these requests. This allows us to fine-tune the execution of these procedures
and handle all requests transparently to the user. However, this approach also has drawbacks. Adding new functionality means adding to a long list of 
Web-service calls and requires substantial implementation effort. In the case of threshold queries the stored procedure performing the evaluation must have
an implementation for each derived field of interest even though the execution is handled by the same Web-service call.

In the future we plan to develop
declarative and graphical user interfaces that will allow users to combine existing building blocks and perform computations that have not been explicitly
implemented. Additionally, we plan on deploying a server-side computing environment for users similar to the CasJobs service for the Sloan Digital Sky Survery \cite{LiThakar}. In such an environment users can run queries in batch mode and save their results in a personal database called MyDB, which 
will allow for much greater flexibility in the type of computations that can be performed in addition to substantially decreasing the network overhead.

@@

\section{Acknowledgment}

The authors would like to thank the Turbulence Database Group at Johns Hopkins University for
their insightful comments and suggestions. This work is supported in part by the National Science Foundation
under Grants ....

>>>>

\begin{abstract}
%Data-intensive architectures and compute clusters have been deployed to store, manage and provide public access to such datasets. 
%The systems built on top of these architectures, sometimes called virtual laboratories, usually do not support the analysis functionality that scientists are 
%interested in out of the box. Moreover, d
In this paper we present a method for the efficient evaluation of threshold queries of derived fields for large numerical 
simulation datasets stored in a cluster of relational databases. 
Such simulations are commonplace in the study of many different physical phenomena. 
The datasets produced by these simulations are in the TB and even PB ranges. 
Data-intensive computations that examine entire time-steps of the simulation data are impractical to perform
locally and can take days or months to iterate over the entire dataset. 
The integrated method for the evaluation of threshold queries that we have developed achieves scalability through data-parallel execution
of the computations on the nodes of the database cluster. 
We extend the database system with the introduction of an application-aware cache for query results, building on the concept of semantic caching.
The cache has little overhead and improves query performance by over an order of magnitude for queries that hit the cache.
\end{abstract}

@@

%Why hasn't it been solved before? (Or, what's wrong with previous proposed solutions? How does mine differ?)
Data-intensive architectures and compute clusters built from commodity hardware rely on parallel I/O to multiple disks and high network bandwidth
to achieve high throughput. Such systems have only recently been deployed for the storage of large numerical simulation 
datasets. The virtual laboratories built on these systems make use of relational database system technology to store and manage these large array datasets.
Relational database systems however often do not support
all of the functionality that scientists are interested in out of the box. It is either up to the user to develop more sophisticated analysis routines locally or such
capabilities have to be built-in through user-defined functions or stored procedures.

Database systems support rollup queries, including top-$k$ queries, but in most cases these
queries apply only simple linear score functions on the attribute values of individual records \cite{Ilyas}.
Additionally, many top-$k$ query evaluation techniques rely on the score functions being monotone in order to perform early pruning (see \cite{Ilyas} for a
survey of top-$k$ evaluation strategies).
This is an assumption that we cannot make for the functions used to compute all the different possible derived fields of interest in fluid mechanics.
Even approaches that aim to work with general score functions \cite{Deshpande, Xin} assume that the function operates on the attributes of a single record.
In contrast our approach computes the value of a derived field at each location in the array by examining the data at all neighboring locations, which
are within the kernel of the computation. 
Finally, a top-$k$ approach may not be suitable in the cases where scientists are interested in performing threshold queries at different time-steps as the
same threshold level will produce different number of points in the result set for different time-steps.

@@

@@

The JHTDB provide a variety of data-intensive analysis routines that are executed on the database nodes. These include interpolation, 
differentiation, particle tracking and spatial filtering. These tasks are often data-intensive and in order to leverage the capabilities of the cluster we
have developed data-driven batch processing techniques for their evaluation \cite{KanovSC11, KanovSC12}. What differentiates threshold and
top-$k$ queries from these tasks is the fact that they usually operate on small subsets of the space or a collection of individual target locations, whereas
threshold and top-$k$ queries usually have to examine the entire data volume of a time-step or a signification portion of it. Furthermore, the data product 
of threshold and top-$k$ 
queries is much smaller in relation to the amount of data that need to be examined. This fact combined with the fact that subsequent queries can reuse 
previously computed results makes the query results suitable to caching.

@@

{\bf Derived fields computation:} The databases store only the raw field data from the simulation (e.g. velocity, pressure, magnetic field etc.). However,
the threshold queries of most interest to science users produce all locations where the values of a \emph{derived} field are above a given threshold. Thus, the
derived field in question has to be computed from the raw data first. For most derived fields of interest, this computation has local support. It has an 
associated localized kernel of computation around each grid node. Therefore, the value of the derived field at each grid node depends on the value of the
stored field at all of the grid locations, which are part of the kernel of computation.

@@

Caching the query results
preserves the computational effort in addition to substantially reducing I/O. The cached data are for the particular derived field that was queried
and not the raw data of the simulation fields.
Thus, we do not have to derive the requested field from the raw data for 
queries that hit the cache. This results in a substantial improvement in query performance as we only have to scan
a small set of data and do not need to perform any additional computation.

@@

So far we have presented the effectiveness of evaluating threshold queries of derived fields on the database cluster storing the raw simulation data. The 
data-parallel computation of the derived fields allows us to evaluate a threshold query over
an entire $1024^3$ time-step part of a 20TB dataset in less than two minutes. The introduction of an application-aware cache for the query results of these
queries reduces this time to several seconds when there is a cache hit. In contrast, one of our science collaborators reported that his evaluation of this
functionality performed locally would take over 20 hours to complete. 
To perform the evaluation locally the user requests the derived field of interest from the database by submitting multiple queries over
subregions of a time-step. This is necessary as requesting a derived field over an entire time-step will overload the network. Derived fields may have even
more components than the scalar or vector field stored in the database. For example, the velocity gradient (needed for the computation of the vorticity) has 9
components compared with the 3 components of the velocity. Given a single-precision floating-point representation, this makes the velocity gradient of an
entire time-step at least 36GB in size. A Web-service request will be much larger due to the overhead of wrapping the data in an xml format. After the
field of interest is obtained locally the user has to threshold it to get the final result, which is reasonably fast, but discards most of the data that have been
requested to yield a small in size result.

\subsection{Evaluation of Additional Fields}
The data-parallel evaluation of threshold queries shows stable execution time for different derived fields in addition to the different threshold levels and
time-steps queried. The execution times depend on the complexity of the computation needed to evaluate the particular derived field requested. Figures 
\ref{fig:vorticity_breakdown}, \ref{fig:q_breakdown}, \ref{fig:magnetic_breakdown} show a breakdown of the execution time of threshold queries of different
derived fields, which are evaluated from the raw data and a cold cache using 4 processes per node on a 4 node cluster. Almost the entire time is spent
performing the I/O and computation associated with the derived field requested.

The vorticity field and the Q-criterion have similar I/O requirements as they
have the same kernels of computation and both can be derived from the velocity gradient. The vorticity has 3 components and its computation only examines
6 of the 9 components of the velocity gradient, which are also examined in pairs (see Eq. \ref{eq:curl}). On the other hand, even though the Q-criterion is a
scalar value, it is computed through a non-linear combination of all 9 of the components of the velocity gradient. This means that the velocity gradient has
to be computed at each grid location before the Q-criterion is evaluated, which is reflected in the increased computation time that we observer for the
Q-criterion. The magnetic field is one of the raw fields of the magnetohydrodynamics dataset that are stored in the database. Therefore, there is no
additional computation needed to derive it from the data, every data point has to be simply compared with the threshold level specified. This is why the 
computation time is much smaller compared to the queries for the vorticity and the Q-criterion. The I/O time for the magnetic field is also smaller. This is
because its kernel of computation is a single point and therefore there are no additional data along the boundary that have to be requested from adjacent 
nodes. In that case all of the data needed are available locally for each database node.

In all of these cases the time taken to interrogate the cache is negligible. The mediator time to distribute the queries and assemble the results as well as the
time to transfer them to the user are also substantially smaller than the I/O and computation times. As expected they increase proportionally to the number
of points in the result set.

@@

\section{Related Work}
Only select few database systems offer support for arrays as first-class citizens. Even fewer provide the fault-tolerance, scalability and availability guarantees 
necessary for a system managing multi-terabyte datasets in a production setting. This is part of the reason why we have chosen to represent the array simulation
data in the JHTDB as a collection of binary large objects in a standard relational DBMS and perform the array manipulation tasks necessary at the application level.
The systems that provide
support for arrays and aim to handle large array data efficiently are RasDaMan \cite{Rasdaman}, SciDB \cite{Scidb} and MonetDB/SciQL \cite{Sciql}. RasDaMan
partitions raster objects into tiles, which are stored in a traditional relational database system. This approach is similar to how the numerical simulation data
are handled in the JHTDB. RasDaMan provides RasQL \cite{Rasql}, which is a SQL-92 based query language for the manipulation of raster images. SciDB
is an array database system build from the ground up. Array attributes are partitioned vertically and each attribute array is decomposed into overlapping
chunks. SciDB provides a declarative Array Query Language (AQL) and an Array Functional Language (AFL). Users can create arrays with named
dimensions with AQL and make use of the functional operators defined in AFL, such
as SLICE, SUBSAMPLE, SJOIN, FILTER and APPLY. SciQL's focus is on language design and the integration with SQL:2003 syntax and semantics. It is implemented
within the MonetDB framework \cite{MonetDB}.

@@

The Web-services approach to archived numerical simulation datasets provides public access to high quality simulation data to anyone with an internet
connection. The Web-services methods can be called from any modern programming language and we provide C, Fortran and Matlab client libraries for the
JHTDB. The evaluation of each query submitted through a Web-service call is carried out on the nodes of the database cluster by means of a stored
procedure
or a user-defined function that has been implemented and deployed to handle these requests. This allows us to fine-tune the execution of these procedures
and handle all requests transparently to the user. However, this approach also has drawbacks. Adding new functionality means adding to a long list of 
Web-service calls and requires substantial implementation effort. In the case of threshold queries the stored procedure performing the evaluation must have
an implementation for each derived field of interest even though the execution is handled by the same Web-service call.

In the future we plan to develop
declarative and graphical user interfaces that will allow users to combine existing building blocks and perform computations that have not been explicitly
implemented. Additionally, we plan on deploying a server-side computing environment for users similar to the CasJobs service for the Sloan Digital Sky Survery \cite{LiThakar}. In such an environment users can run queries in batch mode and save their results in a personal database called MyDB, which 
resides on the servers near the data.
This will allow for much greater flexibility in the type of computations that can be performed in addition to substantially decreasing the network overhead.

%\section{Acknowledgment}

%The authors would like to thank the Turbulence Database Group at Johns Hopkins University for
%their insightful comments and suggestions. This work is supported in part by the National Science Foundation
%under Grants ....

@@
